---
output: 
  pdf_document:
    keep_tex: true
editor_options: 
  chunk_output_type: console
---

\newcommand\textpkg[1]{\textbf{\texttt{#1}}}

```{r, include = FALSE}
# Demonstration of QTLExperiment and multistateQTL 
# Amelia Dunstone
# 2024-10-02

# Analysis for Chapter 5: Case Study Lung Data
```

```{r knitr-options, echo=FALSE}
knitr::opts_chunk$set(
    # warning=FALSE, 
    # error=FALSE, 
    # message=FALSE, 
    fig.height=6, 
    fig.width=8)
```

```{r setup, message = FALSE, results = 'hide', echo = FALSE}
library(dplyr)
library(here)
library(patchwork)
```

<!-- NOTES: -->
<!-- # section -->
<!-- ## subsection -->
<!-- ### subsubsection -->


This chapter demonstrates the use of the packages \textpkg{QTLExperiment} and \textpkg{multistateQTL} with the single-cell lung tissue data from \textcite{Natri_2024_lung}.
This paper investigated genetic variants which may be contributing to the development of Pulmonary Fibrosis (PF), as well as identifed eQTLs specific to a certain cell-type or associated with disease status.
The data consists of lung cells from 67 donors with interstitial lung disease (ILD) and 49 healthy donors.
In this study, *cis* eQTLs were tested, but with a large searching distance (1 Gb). 

Single-cell RNA sequencing experiments often have hundreds of thousands of cells, and in this particular data set there were 475,047 cells following quality control.
In most multi-state QTL pipelines, it would be too computationally demanding to treat each cell-type as a state.
Therefore to perform the multi-state eQTL analysis, we will use results from the aggregated scRNA-seq data.
Unlike tissue type which is a known clinical variable, cell-types must be determined during the annotation step of scRNA-seq analysis.


# Single-cell RNA-seq analysis

We start from the processed Seurat object, available in GEO. 
This object has undergone various analysis steps using Seurat version 4.0 \autocite{Hao_2021_Seurat}. 
The authors of the \textcite{Natri_2024_lung} paper describe these steps in detail in the Methods, but I will provide an abridged version here. 
The scRNA-seq analysis included data filtering, clustering, cell-type annotation and removal of doublets.
The data were classified into 38 cell-types and 4 lineages: immune, epithelial, mesenchymal, and endothelial cells.

In the analysis in this chapter I have subset the data to 10 cell types from across the 4 lineages. 
These cell-types were selected as they are the most abundant cell-types for each lineage. 
The scripts `5_subset_seurat.R` and `6_explore_seurat.R` show the R code that I used to take a smaller subset of the Seurat object, and plot the data. 
As the data object supplied on *GEO* was processed, the only required steps were computing principal components, obtaining clusters, and plotting the data.
The selected cell-types are shown below, overlayed on a Uniform Manifold Approximation and Projection (UMAP) plot \autocite{McInnes_2018_UMAP}. 
UMAP plots are a type of dimensionality reduction commonly used for visualising scRNA-seq data. 
In this plot, each dot represents a cell, and the two-dimensional representation of the cells aims to reflect the high-dimensional structure of the gene expression data.
 
\begin{figure}[h]
\centering
\includegraphics{Figures/scRNAseq_v2.png}
\caption{UMAP plot of single-cell RNA-sequencing gene expression. Each point represents a cell, and they are coloured by lineage. The cell-type labels are overlayed in their approximate average location. The light grey cells show the cells which are not included in the analysis in this chapter.}
\end{figure}


To perform eQTL analysis using single-cell data, it is common to aggregate the gene expression counts. 
\textcite{Natri_2024_lung} normalised the raw gene expression counts and aggregated the data to get a single value for each gene, donor, and cell-type. 
In this paper, gene expression was aggregated by taking the mean expression for the cell-types. 
Other aggregation methods include taking the sum of the gene expression or the median gene expression, and these options are compared in \autocite{Cuomo_2021_Optimising}.
At this stage, the data can be analysed in the same way as was done for the bulk tissue scenario in chapter \ref{Ch4_GTEx}.
The *cis*-eQTL mapping was performed using LIMIX \autocite{Lippert_2014_LIMIX}, which implements linear mixed models.
The summary statistics output from LIMIX include betas, errors and p-values for each association test (gene ID and variant ID) in each state. 
The cell-types (states) I have included are shown in the below table. 
<!-- SNPs were filtered to include SNPs with an MAF greater than 5% and other filters, overall there were 1.9 million SNPs. Associations were tested 1 Gb upstream and downstream of the gene body (cis eQTLs) -->
<!--   -->


\footnotesize
```{r, echo = FALSE}
input_path <- here("Natri_2024_Lung/data/limix/")

state <- read.table(file.path(input_path, "filenames_limix"))$V1
state <- gsub("_qtl_results_all.txt", "", state)

input <- data.frame(
    state=state, 
    path=paste0(input_path, state, "_qtl_results_all.txt"))
input$broad_type <- sapply(strsplit(input$state, "_"), "[", 1)

# Most abundant cell types (most number of people with at least 5 cell types)

# Immune: MDM, Inflammatory monocyte, moDC,
# 
# Mesenchymal: Advential FB, Alveolar FB, SMC?
#   
# Epithelial: Alveolar type 2, Ciliated, Sec - SCGB1A1+/SCGB3A2,
# 
# Endothelial: Arteriole, Venule, aCap?

# 10 states, 11 minutes to load
selected_states <- c("immune_Monocyte-derivedmacrophage", 
    "immune_Inflammatorymonocyte", 
    "immune_moDC", 
    "mesenchymal_AdventitialFB",
    "mesenchymal_AlveolarFB",
    "epithelial_AT2",
    "epithelial_Ciliated",
    "epithelial_Secretory-SCGB1A1+SCGB3A2+",
    "endothelial_arteriole",
    "endothelial_venule")

input_reduced <- input[input$state %in% selected_states,]

input_table <- input_reduced[, c("broad_type", "state")]    

input_table %>% knitr::kable(col.names = c("Lineage", "Cell-type"))

# Add cell counts
# colData <- read.csv(here("Natri_2024_Lung/data/seurat/colData.csv"))
# input_table$counts <- colData$
# input_table %>% knitr::kable(col.names = c("Lineage", "Cell-type", "Number of cells"))
```
\normalsize

## Downloading and preparing data

The LIMIX summary statistics were downloaded from the GEO link in the 'Data Availability' section of the \textcite{Natri_2024_lung} paper and the `.tar.gz` files were extracted (See `1_download_files.R` and `2_extract_data.sh`).  
The data was then subset to 10 cell-types in `3_subset_data.R`. 


# Joint cell-type eQTL analysis with \texttt{mashr}

This section demonstrates the usage of multivariate adaptive shrinkage \autocite{Urbut_2019_mashr} in R. 
The analysis follows the steps outlined in the \href{https://stephenslab.github.io/mashr/articles/eQTL_outline.html}{`eQTL analysis outline'} vignette from \textcite{Stephens_2023_eQTL}, but is modified to make use of \textpkg{QTLExperiment} functions.
The time intensive code chunks in this section are not evaluated, but the same code is included in the script `4_run_mashr.R`. 

I begin by loading the required packages and reading in the `qtle` object which contains the summary statistics. 

\footnotesize
```{r, message = FALSE, results = 'hide', warning = FALSE}
library(QTLExperiment)
library(multistateQTL)

qtle <- readRDS(file ="Natri_2024_Lung/data/qtle_limix_10_states.rds")
```
\normalsize

## Obtain 'strong' and 'random' subsets

The 'eQTL analysis outline' provides an overview of the steps that can be used to run mashr, tailored for multi-state eQTL data. 
As eQTL data has millions of tests across dozens of conditions, it can be slow to run mashr on the full data set for each step. 
There are some steps which produce unbiased statistical estimates if performed on a random subset, so this technique can be used to speed up analysis. 
\textcite{Stephens_2023_eQTL} proposes a workflow which includes two subsets of tests for the data, referred to as the "strong" subset and the "random" subset. 

- **strong subset**: Results from a subset of "strong" tests corresponding to stronger effects in the study. These are helpful to model data driven covariance matrices. 
- **random subset**: The "random" subset contains an unbiased random subset of all tests. This is used to estimate the signal and null distribution of the data.

We select 5000 random rows from the QTLExperiment object, without replacement. 
\footnotesize
```{r}
# Identify a random subset of 5000 tests
random.subset <- sample(nrow(qtle), 5000)
head(random.subset)
```
\normalsize

Using the `getTopHits()` function from \textpkg{multistateQTL}, we select the global eQTLs to form our strong subset. 
The global eQTL refers to the eQTL for each feature ID that has the most significant minimum p-value.
Hence, each row is a unique feature. 
Prior to this, it is necessary to remove NA values from the data. 
Rows with NA values in more than half of the states are removed. 
The remaining NA values are imputed using default values.
After filtering to the top eQTLs for each gene, there are 9118 genes, each with one variant ID.


\footnotesize
```{r}
# Remove and impute NA values
qtle_na <- getComplete(qtle, n=0.5)
qtle_na <- replaceNAs(qtle_na)

# Take the top test for each gene for the 'strong' subset
qtle_strong <- getTopHits(qtle_na, assay="pvalues", mode="state")

# Convert to just indices
strong.subset <- which(rownames(qtle) %in% rownames(qtle_strong))
length(strong.subset)
```
\normalsize

## Run `mashr`

The code chunk below converts the full QTLExperiment object into a `mash` object. 

\footnotesize
```{r, message = FALSE, results = 'hide', warning = FALSE}
library(ashr)
library(mashr)

# Convert to mash object
mashed <- mash_set_data(Bhat = betas(qtle), Shat = errors(qtle))
```
\normalsize

Next we learn the correlation structure using the random subset.
It is important that the sample used at this stage is unbiased as it needs to include null and non-null tests in similar proportions to the original data.

\footnotesize
```{r}
data.temp <- mash_set_data(
    mashed$Bhat[random.subset,], 
    mashed$Shat[random.subset,])
Vhat <- estimate_null_correlation_simple(data.temp)
rm(data.temp)

data.random <- mash_set_data(
    mashed$Bhat[random.subset,], 
    mashed$Shat[random.subset,], 
    V=Vhat)
data.strong <- mash_set_data(
    mashed$Bhat[strong.subset,], 
    mashed$Shat[strong.subset,], 
    V=Vhat)
```
\normalsize

The functions below are used to learn data-driven covariance matrices from the subset of 'strong' test results. 
The goal of this step is to estimate the relationship between different states based on the observed data.

\footnotesize
```{r, eval = FALSE}
U.pca <- cov_pca(data.strong, 5)
U.ed <- cov_ed(data.strong, U.pca)
```
\normalsize

Next we fit the mashr model to the random tests, to learn the mixture weights on all the different covariance matrices and scaling coefficients.

\footnotesize
```{r, eval = FALSE}
# Fit the mashr model 
U.c <- cov_canonical(data.random)

m <- mash(data.random, Ulist = c(U.ed, U.c), outputlevel = 1)
```
\normalsize

\footnotesize
```{r, echo = FALSE}
m <- readRDS("Natri_2024_Lung/data/mashr_10_states.rds")
```
\normalsize

The below functions are exported from `mashr` and generate the assays which can be loaded into the QTLExperiment object.

- `get_pm`: posterior mean
- `get_psd`: posterior standard deviation
- `get_lfsr`:  local false sign rate

The fitted values can be calculated for any subset, but here we have used the 'strong' subset as in the eQTL analysis tutorial. 

\footnotesize
```{r}
mfit = mash(data.strong, g=get_fitted_g(m), fixg=TRUE)

betas.strong <- get_pm(mfit)
errors.strong <- get_psd(mfit)
lfsrs.strong <- get_lfsr(mfit)
```
\normalsize

The figure below compares the distribution of p-values from the `LIMIX` output with the distribution of LFSRs from the `mashr` output. 
The distribution has been shifted to the right after accounting for the information from other states.
Some significant p-values will not be significant after this correction. 

```{r, include = FALSE}
ylims <- c(0, 150000)
```


\footnotesize
```{r pval-plot, fig.asp = 1/2, fig.cap = "Comparison of distribution of p-values from LIMIX output (left), and distribution of LFSRs from mashr output (right).", message = FALSE}

par(mfrow = c(1,2))
hist(pvalues(qtle_strong), xlab = "p-values", main = NULL, ylim = c(0, 150000))
hist(lfsrs.strong, xlab = "LFSRs from mashr", main = NULL, ylim = c(0, 150000))
```

\normalsize

# Multi-state QTL analysis

We use the `QTLExperiment()` constructor to create a QTLExperiment from the estimates, standard errors, and LFSRs output from `mashr`.
The numeric assays 'betas', 'errors' and 'lfsrs' are specified as a list, and passed in through the `assays` argument. 
\footnotesize
```{r}
qtle <- QTLExperiment(
    assays=list(
        betas=betas.strong,
        errors=errors.strong,
        lfsrs=lfsrs.strong),
    metadata=list(study="Natri_Lung"))
qtle
```
\normalsize

```{r, include = FALSE}
saveRDS(qtle, "Natri_2024_Lung/data/qtle_mashr_10_states.rds")
```

As the basic operations of \textpkg{QTLExperiment} and \textpkg{multistateQTL} have been discussed in Chapter 4, we will briefly perform the necessary filtering and calculations.  
We begin by updating the colData of the object to include lineage and cell-types. 
\footnotesize
```{r}
head(colnames(qtle))
# Split state_id into lineage and cell type (state)
qtle$lineage <- sapply(strsplit(state_id(qtle), "_"), "[", 1)
state_id(qtle) <- sapply(strsplit(state_id(qtle), "_"), "[", 2)
head(colData(qtle))
```
\normalsize

## Plotting global sharing

A main aim of multi-state eQTL analysis is to better understand the relationship between states.
One way to measure this is by the proportion of eQTL effects which are shared between pairs of states. 

We begin by calling significance, using a threshold of 0.05 for the LFSRs. 
We then filter to only rows with some significant states, and calculate the pairwise sharing using `runPairwiseSharing()`.
The function `plotPairwiseSharing()` produces a heatmap showing the proportion of significant eQTLs which are shared for each pairwise combination of states. 
This can help us to understand the relationship between the cell-types by their eQTL results. 
As the columns (and rows) of the plot are clustered, we can see which states are most related. 
For example, the states from the lineage 'mesenchymal' have less sharing than the immune cell-types. 

\footnotesize
```{r plot-pairwise-lung}
qtle <- callSignificance(qtle, assay="lfsrs", thresh=0.05)

qtle_sig <- getSignificant(qtle)
qtle_top <- runPairwiseSharing(qtle_sig)

dim(qtle_top)

plotPairwiseSharing(qtle_top, annotateColsBy=c("nSignificant", "lineage"))
```
\normalsize
\footnotesize
```{r plot-upset-lung, fig.asp = 2/3}
plotUpSet(qtle_top, annotateColsBy=c("nSignificant", "lineage"))
```
\normalsize


## Characterising multi-state QTL patterns

After using `runTestMetrics()` to characterise eQTLs into 'unique', 'multi-state' or 'global', we can see that the majority of eQTLs are unique. 
\footnotesize
```{r}
qtle_top <- runTestMetrics(qtle_top)

table(rowData(qtle_top)$qtl_type)
```
\normalsize
\footnotesize
```{r plot-qtl-clusters-lung, include = FALSE, fig.asp = 2/3}
qtle_top_ms <- subset(qtle_top, qtl_type_simple == "multistate")

# qtle_top_ms <- subset(qtle_top_ms, )
plotQTLClusters(
    qtle_top_ms, 
    annotateColsBy="lineage",
    annotateRowsBy="qtl_type")
```
\normalsize

We can subset to just the unique eQTLs and plot a heatmap of these beta values. 
This shows that some states have unique eQTLs which have strong effects on the gene expression. 
For example, the state 'AdventitialFB' has a number of unique eQTLs which cause an increase in the expression of a gene, as well as eQTLs that result in a decrease in gene expression.

\footnotesize
```{r plot-unique-lung, fig.asp = 0.9}
qtle_top_unique <- subset(qtle_top, qtl_type_simple == "unique")

plotQTLClusters(
    qtle_top_unique, 
    annotateColsBy="lineage",
    annotateRowsBy=c("qtl_type"))
```
\normalsize

# Conclusion

This concludes our demonstration of \textpkg{QTLExperiment} and \textpkg{multistateQTL}. 
In this chapter, we analysed a data set where states represent different cell-types, outlining the additional challenge that comes with analysing single-cell data.
We mentioned the processing steps which are required before the multi-state QTL analysis can begin, including quality control, batch correction, feature selection, clustering, and principal component analysis. 
Notably, it is necessary to annotate the cell-types in the data set so that these can be used as the 'states' in the multi-state eQTL analysis which follows. 

We also provided a tutorial to run multivariate adaptive shrinkage using functions from the \textpkg{multistateQTL} package. 
This adjusts the summary statistics from condition-by-condition results so that patterns of sharing is learnt between states. 
Finally, we visualised the multi-state eQTL data to identify patterns within the states, and gained a deeper understanding of the impact genetic variants have across a range of cell-types. 


```{r, include = FALSE}
knitr::knit_exit()
```

```{r}
sessionInfo()
```