---
title: "Illustration of Adaptive Shrinkage"
author: "Amelia Dunstone"
date: "2024-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(REBayes)

library(ashr)
library(ggplot2)
library(scales)

set.seed(100)
library(here)
# Not working but can run function
source(here("Urbut_Mashr/datamaker.R"))
```

```{r}
NSAMP = 1000
s     = 1/rgamma(NSAMP,5,5)

sim.spiky =
  rnormmix_datamaker(args = list(g = normalmix(c(0.4,0.2,0.2,0.2),
                                               c(0,0,0,0),
                                               c(0.25,0.5,1,2)),
                                  min_pi0   = 0,
                                  max_pi0   = 0,
                                  nsamp     = NSAMP,
                                  betahatsd = s))

sim.bignormal =
  rnormmix_datamaker(args = list(g         = normalmix(1,0,4),
                                 min_pi0   = 0,
                                 max_pi0   = 0,
                                 nsamp     = NSAMP,
                                 betahatsd = s))

cat("Summary of observed beta-hats:\n")


print(rbind(spiky     = quantile(sim.spiky$input$betahat,seq(0,1,0.1)),
            bignormal = quantile(sim.bignormal$input$betahat,seq(0,1,0.1))),
      digits = 3)




beta.spiky.ash     = ash(sim.spiky$input$betahat,s)
beta.bignormal.ash = ash(sim.bignormal$input$betahat,s)


make_df_for_ashplot =
  function (sim1, sim2, ash1, ash2, name1 = "spiky", name2 = "big-normal") {
    n = length(sim1$input$betahat)
    x = c(get_lfsr(ash1),get_lfsr(ash2))
    return(data.frame(betahat  = c(sim1$input$betahat,sim2$input$betahat),
                      beta_est = c(get_pm(ash1),get_pm(ash2)),
                      lfsr     = x,
                      s        = c(sim1$input$sebetahat,sim2$input$sebetahat),
                      scenario = c(rep(name1,n),rep(name2,n)),
                      signif   = x < 0.05))
  }

ashplot = function(df,xlab="Observed beta-hat",ylab="Shrunken beta estimate")
  ggplot(df,aes(x = betahat,y = beta_est,color = 1/s)) +
    xlab(xlab) + ylab(ylab) + geom_point() +
    facet_grid(.~scenario) +
    geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
    scale_colour_gradient2(midpoint = median(1/s),low = "blue",
                           mid = "white",high = "red",space = "Lab") +
    coord_fixed(ratio = 1)
```

Two key features of the plots illustrate the ideas of adaptive shrinkage: i) the estimates under the spiky scenario are shrunk more strongly, illustrating that shrinkage adapts to the underlying distribution of beta; ii) in both cases, estimates with large standard error (blue) are shrunk more than estimates with small standard error (red) illustrating that shrinkage adapts to measurement precision.

```{r}
df = make_df_for_ashplot(sim.spiky,sim.bignormal,beta.spiky.ash,
                         beta.bignormal.ash)
print(ashplot(df))



z_lfsr_plot = function(df,ylab = "Observed Z score",xlab = "lfsr")
  ggplot(df,aes(x = lfsr,y = betahat/s,color = 1/s)) +
    xlab(xlab) + ylab(ylab) + geom_point() + facet_grid(.~scenario) +
    scale_colour_gradient2(midpoint = median(1/s),low = "blue",
                           mid = "white",high = "red",space = "Lab")

print(z_lfsr_plot(df))
```
